---
title: "dawnn"
output:
    rmarkdown::html_vignette:
        fig_width: 5
        fig_height: 3
vignette: >
  %\VignetteIndexEntry{dawnn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = "#>"
)
```

<details>
<summary>We have simulated some toy data to demonstrate Dawnn. Click here to display the simulation code.</summary>
We need to first generate a dataset on which we can run Dawnn. We generate
three clusters of cells. In one cluster, we aim to have a 50/50 split of the
labels "Condition1" and "Condition2"; in the second, we aim for a 10/90 split;
and in the third we aim for a 90/10 split. This means that there is
differential abundance in the second and third clusters. We aim to detect this
with Dawnn.

```{r create_seurat_obj}
library(stats)
library(dplyr)
library(Seurat)

set.seed(123)

# Simulate three samples with the expression of 30 genes measured:
#   - Sample 1 is upregulated in the first ten genes
#   - Sample 2 is upregulated in the second ten genes
#   - Sample 3 is upregulated in the third ten genes
sample_1 <- rnbinom(n = 30000, size = 1,
                    prob = c(rep(0.5, 0), rep(0.1, 10), rep(0.5, 20)))
sample_2 <- rnbinom(n = 30000, size = 1,
                    prob = c(rep(0.5, 10), rep(0.1, 10), rep(0.5, 10)))
sample_3 <- rnbinom(n = 30000, size = 1,
                    prob = c(rep(0.5, 20), rep(0.1, 10), rep(0.5, 0)))

ge_vect <- c(sample_1, sample_2, sample_3)
ge_matrix <- matrix(ge_vect, ncol = 3000)
colnames(ge_matrix) <- paste0("cell", 1:3000)
rownames(ge_matrix) <- paste0("gene", 1:30)

cells <- CreateSeuratObject(counts = ge_matrix) %>%
         NormalizeData() %>%
         FindVariableFeatures() %>%
         ScaleData() %>%
         RunPCA() %>%
         RunUMAP(dims = 1:10)

cells$pc1 <- c(rep(0.1, 1000), rep(0.5, 1000), rep(0.9, 1000))
cells$label <- ifelse(runif(3000) <= cells$pc1, "Condition1", "Condition2")
DimPlot(cells, group.by = "label")
```
</details>

## Main workflow

```{r setup}
library(dawnn)
library(Seurat)
```

The first time we use Dawnn, we need to download its trained neural network
model. For the majority of cases, simply calling `download_model()` will be
sufficient. By default, the model is downloaded to
`~/.dawnn/dawnn_nn_model.h5`.

```{r download_model, eval = FALSE}
download_model()
```

Once we have downloaded the model, we can call Dawnn with ``run_dawnn()``.

```{r run_dawnn}
cells <- run_dawnn(cells, label_names = "label", label_1 = "Condition1",
                   label_2 = "Condition2", reduced_dim = "pca")
```

### Dawnn's non-optional parameters

`run_dawnn` has five parameters that we must specify:

| Parameter | Description |
|-----------|-------------|
| `cells`   | Seurat object containing the dataset |
| `label_names` | `meta.data` slot in `cells` containing labels |
| `label_1` | The label corresponding to the first sample |
| `label_2` | The label corresponding to the second sample |
| `reduced_dim` | Dimensionality reduction to use when calculating KNN graph |

### Dawnn's outputs

Dawnn's outputs are stored in `meta.data` slots of `cells`. These outputs are:

| Dawnn output             | Description                                                                                   |
|--------------------------|-----------------------------------------------------------------------------------------------|
| `cells$dawnn_da_verdict` | Boolean output of Dawnn for whether it is in a region of differential abundance.              |
| `cells$dawnn_lfc`        | Estimated log2-fold change in its neighbourhood.                                              |
| `cells$dawnn_scores`     | Estimated probability that cell drawn from sample associated with label_1.                    |
| `cells$dawnn_p_vals`     | P-value associated with the hypothesis test that it is in a region of differential abundance. |

The first two outputs are likely the most useful.  `dawnn_da_verdict` tells us
whether Dawnn has called a cell as being in a region of differential abundance
and `dawnn_lfc` contains the estimated log2-fold change in the abundance of
`label_1` and `label_2` in the neighbourhood of each cell.

```{r plot_first_two_dawnn_results}
DimPlot(cells, group.by = "dawnn_da_verdict")
FeaturePlot(cells, "dawnn_lfc")
```

We can extract the cells for which `cells$dawnn_da_verdict` is `TRUE` in order
to find the subpopulations perturbed in the experiment

`dawnn_scores` contains the direct output of Dawnn's neural
network, i.e. the estimated probability a cell was drawn from the sample associated
with `label_1`. `dawnn_scores` is converted into `dawnn_lfc` using
`log2(cells$dawnn_scores / (1 - cells$dawnn_scores))`. Finally, `dawnn_p_vals`
contains, for each cell, the p-value associated with testing the null
hypothesis "this cell is not in a region of differential abundance".

```{r plot_second_two_dawnn_results}
FeaturePlot(cells, "dawnn_scores")
FeaturePlot(cells, "dawnn_p_vals")
```

## Optional parameters

### Setting all options

In this example, we change all options that can be passed to Dawnn.

```{r dawnn_all_options, eval = FALSE}
cells <- run_dawnn(cells, label_names = "labels", label_1 = "Condition1",
                   label_2 = "Condition2", reduced_dim = "pca", n_dims = 20,
                   nn_model = "~/another_dawnn_model.h5",
                   recalculate_graph = FALSE, alpha = 0.05, verbosity = 0,
                   seed = 42)
```

### Changing model location

We can use `download_model`'s `model_file_path` parameter to change the
location to which the model is downloaded. This location can be passed to
`run_dawnn` using the `nn_model` parameter.

```{r changing_model_location, eval = FALSE}
download_model(..., model_file_path = "~/Documents/new_model_location.h5")
run_dawnn(..., nn_model = "~/Documents/new_model_location.h5")
```

### Downloading a model from another location

A model can be downloaded from a non-default url using the `model_url`
parameter.

```{r download_model_another_url, eval = FALSE}
download_model(..., model_url = "example.com/another_model_url.h5")
```
